---
Purpose: "12709 - Data Analytics Project"
Title: "Analysis, Modeling and Predictions of Fuel Economy Dataset"
author: "Simran Bhalla, Tanmay Nema, Gautam Petkar"
date: "12/8/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Title: "Analysis, Modeling and Predictions of Fuel Economy Dataset"
### "Simran Bhalla, Tanmay Nema, Gautam Petkar"

## Introduction

_### Objective_
The objective of this analysis is to explore the underlying trends and track the changes in the transportation sector for passenger vehicles, with specific emphasis on fuel efficiency and carbon emissions.

_### Dataset_
The dataset chosen for the project has been taken from [fueleconomy.gov](https://www.fueleconomy.gov/feg/ws/index.shtml#vehicle) which is the official U.S. government source for fuel economy information. The website helps consumers make informed fuel economy choices before choosing a vehicle to purchase and helps them achieve the best fuel economy possible from the cars they own. This website is maintained by the U.S. Department of Energy’s (DOE’s) Office of Energy Efficiency and Renewable Energy with data being provided by the U.S. Environment Protection Agency (EPA). The website also helps in fulfilling the DOE’s and EPA's responsibility under the Energy Policy Act of 1992 to provide accurate fuel economy information to consumers. On the website it is possible to find and compare cars by model, make or class and it also shows the best and worst vehicles as well as the most viewed vehicles daily.  

The purpose of EPA’s fuel economy estimates is to provide a reliable basis for comparing vehicles. Some of the key parameters that the dataset consists are miles-per-gallon information for different use-case such as city application, highway application and combined application (55% city – 45% highway), tailpipe CO2 emissions, fuel consumption, cost savings, etc. Estimates for all vehicles are based on laboratory testing under standardized conditions to allow for fair comparisons. The vehicle classes used for comparison are two-seater cars, sedans, station wagons, pickup trucks, vans, minivans, SUVs, etc.  

For this project, the vehicle dataset which is available on the website is used. It contains fuel economy information for 1984-current model year vehicles. The data is available for download in CSV and XML formats.  

_### Data Dictionary_
The data dictionary was available on the website and is shown in Appendix-A.  

_### Choice of Tool Used_ 
To be able to successfully complete our analysis we make use of Microsoft Excel and R. The dataset is available in .csv format and we perform an initial exploratory data analysis using Microsoft Excel. This helps us understand the variables we want to use and gives us an initial idea of the research questions we can propose. Excel makes it easier to look at the data at one go and add filters to the variables. By filtering the data, we also get an idea of null values which further assist in the process of data cleaning.  
After looking at the data using Microsoft Excel, we perform our Exploratory Data Analysis and Data Manipulation using R programming language on RStudio software package. The summary() and the str () function make it easy to look at the type of data in the variable as well as the minimum, maximum, median and mean values of the variables in the dataset. We went ahead with R for the bulk of our analysis since it offers the flexibility of installing additional packages to carry out advanced analysis and modeling using statistical and machine learning models.  
Further, we used git interfaced with RStudio to facilitate version control and enable coding collaboration between the group members.  
For the statistical analysis and modelling of our research questions we use the tools and packages in R such as - *add tools and used for each research question*.

```{r}
# Libraries used
library(tidyverse)
library(dplyr)
library(ggplot2)
library(gridExtra)
library(stringr)
library(ggpubr)
library(rstatix)
library(forecast)
library(lmtest)
```

## Exploratory Data Analysis - Data Cleaning - Data Manipulation

The structure and summary of the raw dataset is displayed below:

```{r}
# Import data from csv and save in a data frame
holder <- read.csv("vehicles.csv")
raw_data_vehicle <- tibble(holder)

#Dimesion of the raw dataset:
dim(raw_data_vehicle)
```

```{r}
##------------------------------------------------------------------------------

# Calculation of volume of vehicle (parameter used in further analysis)
vols <- raw_data_vehicle %>%
  select(hpv,lv4, pv4, pv2,lv2, hlv)
vols <- rowSums(vols)

#removing unwanted columns
data <- raw_data_vehicle %>%
  select(Year = year, Make = make, Model =model, FuelType1 = fuelType1, FuelType2 = fuelType2, CityMPG = city08, HighMPG = highway08, CombinedMPG = comb08, C02 = co2TailpipeGpm, SuperCharger = sCharger, TurboCharger = tCharger,  FuelConsumption = barrels08, Cylinders = cylinders, Class = VClass, EngineDisplacement = displ) %>%
  mutate(VehicleVolume = vols)
##------------------------------------------------------------------------------
# Convert Categorical variables to Factors
data <- data %>%
  mutate(Make = as.factor(Make), Model = as.factor(Model), FuelType1 = as.factor(FuelType1), FuelType2 = as.factor(FuelType2), SuperCharger = as.factor(SuperCharger), TurboCharger = as.factor(TurboCharger))

# Replace NA's from data
data$Cylinders <- data$Cylinders %>% replace_na(0)

summary(data)
str(data)
##------------------------------------------------------------------------------
# Create Frequency Tables for EDA
df1 <- as.data.frame(table(data$Make)) %>%
  arrange(desc(Freq))
df2 <- as.data.frame(table(data$FuelType1)) %>%
  arrange(desc(Freq))
df3 <- as.data.frame(table(data$FuelType2)) %>%
  arrange(desc(Freq))
df4 <- as.data.frame(table(data$SuperCharger)) %>%
  arrange(desc(Freq))
df5 <- as.data.frame(table(data$TurboCharger)) %>%
  arrange(desc(Freq))
df6 <- as.data.frame(table(data$Cylinders)) %>%
  arrange(desc(Freq))
##------------------------------------------------------------------------------
# Boxplots for numerical variables for EDA
boxplot(data$CombinedMPG, data$CityMPG, data$HighMPG,
        main = "Spread of MPG attributes",
        at = c(1,3,5),
        names = c("Combined MPG", "City MPG", "Highway MPG"),
        col = c("orange","red"),
        border = "brown",
        horizontal = TRUE,
        notch = TRUE)

boxplot(data$C02,
        main = "Spread of Tailpipe Emission (g/mile)",
        at = c(1),
        names = c("CO2"),
        col = c("orange","red"),
        border = "brown",
        horizontal = TRUE,
        notch = TRUE)

boxplot(data$FuelConsumption,
        main = "Spread of Fuel Consumption (Barrels)",
        at = c(1),
        names = c("Fuel Consumption"),
        col = c("orange","red"),
        border = "brown",
        horizontal = TRUE,
        notch = TRUE)
##------------------------------------------------------------------------------

## Data Cleaning: 

# Combining all the Gasoline Types as one 
clean_data <- data %>%
  mutate(FuelType1 = str_replace_all(string = FuelType1, pattern = "Regular Gasoline", replacement = "Gasoline")) %>%
  mutate(FuelType1 = str_replace_all(string = FuelType1, pattern = "Premium Gasoline", replacement = "Gasoline")) %>%
  mutate(FuelType1 = str_replace_all(string = FuelType1, pattern = "Midgrade Gasoline", replacement = "Gasoline"))

#Combining class
clean_data <- clean_data %>% 
  mutate(Class = case_when(str_detect(Class, pattern = "Trucks") ~ "Pickup Trucks",
                              str_detect(Class, pattern = "Cars|Wagons") ~ "Cars",
                              str_detect(Class, pattern = "Sport") ~ "SUV",
                              str_detect(Class, pattern = regex("van", ignore_case = TRUE) ) ~ "Vans", 
                              TRUE ~ "Other"))
##------------------------------------------------------------------------------

# Comparison Before and After cleaning
c1 <- unique(data$FuelType1)
c2 <- unique(clean_data$FuelType1)
comp1 <- data.frame(Unclean = c1, Clean = c(c2,NA,NA))

print("Comparison of Data Before and After cleaning")
comp1
str(clean_data)
summary(clean_data)

#Specifying Categorical Variable as Factor
clean_data$FuelType1 <- as.factor(clean_data$FuelType1)

##------------------------------------------------------------------------------
```


```{r}
# Other EDA charts

##------------------------------------------------------------------------------

# Vehicles and Manufacturers
clean_data %>% 
  count(Make, Model) %>% count(Make) %>% top_n(20, wt = n) %>% 
  ggplot(aes(x = reorder(Make, n), y = n )) +
  geom_bar( aes(fill = Make), stat = "identity", show.legend = FALSE) + 
    labs(title = "Top manufacturers by number of models", y = "Count of models", x = "Make") +
  coord_flip() 

##------------------------------------------------------------------------------

# Fuel Economy over the years by fuel type
clean_data %>% 
  group_by(Year, FuelType1) %>% 
  summarise(avgMPG = mean(CombinedMPG)) %>% 
  ggplot(aes(x=Year, y = avgMPG)) +
  geom_line(aes(color = avgMPG), alpha = 0.7, show.legend = FALSE) +
  geom_point(aes(color = avgMPG), alpha = 0.5, show.legend = FALSE) +
  scale_x_continuous(breaks =  seq( min(clean_data$Year), max(clean_data$Year), by = 2) ) +
  labs(title = "Average Fuel Economy by Fuel Type", y = "Avg MPG", x = "Year") +
  facet_wrap(~ FuelType1, scales = "free") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1), legend.direction = "horizontal", legend.position = "top")
```

## Research Questions

### Question 1: Are vehicle with Supercharger more fuel efficient than vehicle with Turbochargers?
### (Inferential question - T-test)
```{r}
vehicles_SCTC <- clean_data %>% 
  filter(SuperCharger == 'S' | TurboCharger == 'TRUE') %>%
  mutate(Charger = if_else(SuperCharger == 'S','SC','TC',missing = NULL))

vehicles_SCTC %>%
  group_by(TurboCharger, SuperCharger) %>%
  summarise(n = n(), mean = mean(CombinedMPG), median = median(CombinedMPG), sd = sd(CombinedMPG))

tc <- vehicles_SCTC %>%
  filter(TurboCharger == 'TRUE' & !SuperCharger == 'S')

sc <- vehicles_SCTC %>%
  filter(SuperCharger == 'S' & is.na(TurboCharger))

tcsc <- vehicles_SCTC %>%
  filter(SuperCharger == 'S' & TurboCharger == 'TRUE')

boxplot(tc$CombinedMPG, sc$CombinedMPG, tcsc$CombinedMPG,
main = "Comparison Between Turbocharger and Supercharger Vehicles",
at = c(1,3,5),
names = c("Turbocharger", "Supercharger", "Dual"),
las = 2,
col = c("orange","red"),
border = "brown",
horizontal = TRUE,
notch = TRUE
)

# To verify data is normally distributed
ggplot(vehicles_SCTC, aes(x=CombinedMPG)) +
    geom_histogram(binwidth=2, colour="black", fill="white")

# T-Test on TC/SC set:
stat.test <- t.test(vehicles_SCTC$CombinedMPG ~ vehicles_SCTC$Charger, var.equal = FALSE) %>% add_significance()
stat.test
```
### Question 2: Is there a difference in fuel economy b/w different vehicle classes?
### (Inferential question - ANOVA)
```{r}

clean_data %>%
  distinct(Class)

vehicles_class <- clean_data %>% 
  filter(!is.na(Class)) %>%
  filter(FuelType1 == 'Gasoline' , FuelType2 != "Electricity")

vehicles_class %>%
  group_by(Class) %>%
  summarise(n = n(), mean = mean(CombinedMPG), median = median(CombinedMPG), sd = sd(CombinedMPG))

boxplot(vehicles_class$CombinedMPG ~ vehicles_class$Class,
main = "Comparison Between Vehicle Class",
las = 2,
col = c("orange","red"),
border = "brown",
horizontal = TRUE,
notch = TRUE,
xlab = '',
ylab = '')

# To verify data is normally distributed
ggplot(vehicles_class, aes(x=CombinedMPG)) +
    geom_histogram(binwidth=2, colour="black", fill="white")

# ANOVA Statistical Test
fit_class <- lm(CombinedMPG ~ as.factor(Class), data = vehicles_class)
anova(fit_class)

```
### Question 3: Is there a correlation between Engine Displacement and Vehicle Volume?
### (Inferential question - Correlation)
```{r}

correlation_data <- clean_data %>% 
  filter(FuelType1 != 'Electricity') %>%
  filter(!is.na(EngineDisplacement)) %>%
  filter(VehicleVolume > 20 & VehicleVolume < 230)

correlation_data %>%
  summarise(n = n(), mean = mean(VehicleVolume), median = median(VehicleVolume), sd = sd(VehicleVolume))

boxplot(correlation_data$VehicleVolume,
main = "Spread of Vehicle Volume (cubic ft)",
las = 1,
col = c("orange","red"),
border = "brown",
horizontal = TRUE,
notch = TRUE
)

# To verify data is normally distributed
ggplot(correlation_data, aes(x=EngineDisplacement)) +
    geom_histogram(binwidth=0.2, colour="black", fill="white")

ggplot(correlation_data, aes(x=VehicleVolume)) +
    geom_histogram(binwidth=2, colour="black", fill="white")

# Correlation Test - (not normally distributed data)
res_corr <- cor.test(correlation_data$EngineDisplacement, correlation_data$VehicleVolume, method = "kendall")
res_corr

ggscatter(correlation_data, x = "EngineDisplacement", y = "VehicleVolume", color = "VehicleVolume",
          title = "Correlation between Engine Displacement and Vehicle Volume",
          add = "reg.line", conf.int = TRUE, 
          cor.coef = FALSE, cor.method = "kendall",
          xlab = "Engine Displacement", ylab = "Vehicle Volume",
          add.params = list(color = "red", fill = "lightgray")) + 
          stat_cor(method = "kendall", label.x = 3, label.y = 30)

```

### Question 4: What is the trend of CO2 emission by year for top 5 vehicle manufacturers?
### (Predictive question - Correlation)
```{r}
new_df <- data.frame(vehicles.comb08 = raw_data_vehicle$comb08, vehicles.co2TailpipeGpm = raw_data_vehicle$co2TailpipeGpm, vehicles.year = raw_data_vehicle$year, vehicles.fuelType1 = raw_data_vehicle$fuelType1, vehicles.make = raw_data_vehicle$make)

#Consider only IC engine vehicles
IC_veh <- filter(new_df, vehicles.fuelType1 != "Electricity")

#Consider Companies with highest vehicle count (Top 5)

#Chevy
chevy <- filter(IC_veh, vehicles.make == "Chevrolet")
chevy_avg_co2 <- aggregate(cbind(chevy$vehicles.co2TailpipeGpm)~chevy$vehicles.year, FUN = mean)
ts_chevy_co2 <- ts(chevy_avg_co2$V1, frequency = 1, start = 1984)
autoarima_chevy_co2 <- auto.arima(ts_chevy_co2)
chevy_forecast_co2 <- forecast(autoarima_chevy_co2, h=10)
chevy_co2_forecast <- plot(chevy_forecast_co2, main = "Tailpipe CO2 Emissions Forecast for Chevrolet", xlab = "Years", ylab = "Avg. CO2 Emissions (g/miles)")

#Ford
ford <- filter(IC_veh, vehicles.make == "Ford")
ford_avg_co2 <- aggregate(cbind(ford$vehicles.co2TailpipeGpm)~ford$vehicles.year, FUN = mean)
ts_ford_co2 <- ts(ford_avg_co2$V1, frequency = 1, start = 1984)
autoarima_ford_co2 <- auto.arima(ts_ford_co2)
ford_forecast_co2 <- forecast(autoarima_ford_co2, h=10)
ford_co2_forecast <- plot(ford_forecast_co2, main = "Tailpipe CO2 Emissions Forecast for Ford", xlab = "Years", ylab = "Avg. CO2 Emissions (g/miles)")

#Dodge
dodge <- filter(IC_veh, vehicles.make == "Dodge")
dodge_avg_co2 <- aggregate(cbind(dodge$vehicles.co2TailpipeGpm)~dodge$vehicles.year, FUN = mean)
ts_dodge_co2 <- ts(dodge_avg_co2$V1, frequency = 1, start = 1984)
autoarima_dodge_co2 <- auto.arima(ts_dodge_co2)
dodge_forecast_co2 <- forecast(autoarima_dodge_co2, h=10)
dodge_co2_forecast <- plot(dodge_forecast_co2, main = "Tailpipe CO2 Emissions Forecast for Dodge", xlab = "Years", ylab = "Avg. CO2 Emissions (g/miles)")

#GMC
gmc <- filter(IC_veh, vehicles.make == "GMC")
gmc_avg_co2 <- aggregate(cbind(gmc$vehicles.co2TailpipeGpm)~gmc$vehicles.year, FUN = mean)
ts_gmc_co2 <- ts(gmc_avg_co2$V1, frequency = 1, start = 1984)
autoarima_gmc_co2 <- auto.arima(ts_gmc_co2)
gmc_forecast_co2 <- forecast(autoarima_gmc_co2, h=10)
gmc_co2_forecast <- plot(gmc_forecast_co2, main = "Tailpipe CO2 Emissions Forecast for GMC", xlab = "Years", ylab = "Avg. CO2 Emissions (g/miles)")

#Toyota
toyota <- filter(IC_veh, vehicles.make == "Toyota")
toyota_avg_co2 <- aggregate(cbind(toyota$vehicles.co2TailpipeGpm)~toyota$vehicles.year, FUN = mean)
ts_toyota_co2 <- ts(toyota_avg_co2$V1, frequency = 1, start = 1984)
autoarima_toyota_co2 <- auto.arima(ts_toyota_co2)
toyota_forecast_co2 <- forecast(autoarima_toyota_co2, h=10)
toyota_co2_forecast <- plot(toyota_forecast_co2, main = "Tailpipe CO2 Emissions Forecast for Toyota", xlab = "Years", ylab = "Avg. CO2 Emissions (g/miles)")
```
### Question 5: Is there a difference in the City MPG and Highway MPG values for Gasoline vs EVs?
### (Inferential question - F-test, T-test)
```{r}
#F-Test and T-test on the ratio of City MPG and Highway MPG 
#http://www.sthda.com/english/wiki/unpaired-two-samples-t-test-in-r

gasoline_mpg <- filter(clean_data, FuelType1 == "Gasoline", Make == "Toyota" )
gasoline_mpg$ratio <- gasoline_mpg$CityMPG / gasoline_mpg$HighMPG
EV_mpg <- filter(clean_data, FuelType1 == "Electricity", Make == "Tesla")
EV_mpg$ratio <- EV_mpg$CityMPG / EV_mpg$HighMPG
gas_ratio <- gasoline_mpg$ratio
gas_ratio <- sample(gas_ratio,100,replace=FALSE)
EV_ratio <- EV_mpg$ratio
EV_ratio <- head(EV_ratio, 100)

# Create a new data frame
my_data <- data.frame( 
                group = rep(c("gas_ratio", "EV_ratio"), each = 100),
                ratio = c(gas_ratio,  EV_ratio)
                )

# Summary stats
group_by(my_data, group) %>%
  summarise(
    count = n(),
    mean = mean(ratio, na.rm = TRUE),
    sd = sd(ratio, na.rm = TRUE)
  )

ggboxplot(my_data, x = "group", y = "ratio", 
          color = "Brown", fill = "Orange",
          title = "City vs Highway MPG Ratio for Gasoline vs EVs",
          orientation = "horizontal",
          ylab = "Ratio", xlab = "Groups",
          )

# Normality test
# Shapiro-Wilk normality test for Gasoline MPG ratio
with(my_data, shapiro.test(ratio[group == "gas_ratio"]))# p = 0.0001937

# Shapiro-Wilk normality test for EV MPG ratio
with(my_data, shapiro.test(ratio[group == "EV_ratio"])) # p = 0.0001738

# F test - do they have same variances?
res.ftest <- var.test(ratio ~ group, data = my_data)
res.ftest

# T - test
res <- t.test(ratio ~ group, data = my_data, var.equal = TRUE)
res
```
### Question 6: What is the trend of new vehicle models each year by fuel type?
### (Inferential question - Regression)
```{r}
type_data <- clean_data %>%
  select(Year, FuelType1, FuelType2)

type_data <- type_data %>%
  mutate(VehicleType = ifelse((FuelType1 == 'Gasoline' & is.na(FuelType2)), 'Gasoline', ifelse((FuelType1 == 'Gasoline' & FuelType2 == 'Electricity'), 'HEV', ifelse(FuelType1 == 'Electricity', 'BEV', ifelse((FuelType1 == 'Gasoline' & (FuelType2 == 'E85' | FuelType2 == 'Propane' | FuelType2 == 'Natural Gas')),'Hybrid',FuelType1)))))
 
type_plot <- type_data %>%
  group_by(Year, VehicleType) %>%
  tally()

type_plot <- type_plot %>%
  rename(Count = n) %>%
  filter(Year<2021)

ggplot(type_plot, aes(x = Year, y = Count, color = VehicleType)) + 
  geom_line() + 
  scale_y_log10() +
  ggtitle('Count of VehicleType Trend with Model Year') +
  ylab('Count of VehicleType') + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        axis.title.x = element_blank())

# Regression 
ggplot(data = gasoline, aes(x = Year, y = Count)) +
geom_point() +
stat_smooth(method = "lm", col = "dodgerblue3") +
theme(panel.background = element_rect(fill = "white"),
axis.line.x=element_line(),
axis.line.y=element_line()) +
ggtitle("Linear Model Fitted to Data")

ggplot(data = bev, aes(x = Year, y = Count)) +
geom_point() +
stat_smooth(method = "lm", col = "dodgerblue3") +
theme(panel.background = element_rect(fill = "white"),
axis.line.x=element_line(),
axis.line.y=element_line()) +
ggtitle("Linear Model Fitted to Data")

ggplot(data = hev, aes(x = Year, y = Count)) +
geom_point() +
stat_smooth(method = "lm", col = "dodgerblue3") +
theme(panel.background = element_rect(fill = "white"),
axis.line.x=element_line(),
axis.line.y=element_line()) +
ggtitle("Linear Model Fitted to Data")

```

